

## Reading data file(s)

`DATA_FILES` is read with [edgeR] [@Robinson2010bis] functions to construct the `DEGlist` object using the user-defined function `LoadExpressionData()`.

```{r buildx}
# use 'x' as a short name since it will be thoroughly used
x <- LoadExpressionData(DATA_FILES, DATA_DIR, EXP_FACTORS, COLUMNS_TO_READ)

# raw data will be saved only when the expression matrix was not the input file
SAVE_RAW_DATA <- (length(DATA_FILES) > 1)
```


## Assigning colours to factors

Based on `EXP_FACTORS`, the number of different experimental factors will be extracted to define specific colours for them throughout the report.

```{r FactorsColours}
# convert experimental factors to colours
UNIQ_FACTORS <- unique(EXP_FACTORS)
EXP_COLORS <- as.vector(EXP_FACTORS)
i <- 1
for (i in 1:length(UNIQ_FACTORS)) {
   # replace each factor by a colour number
   EXP_COLORS <- replace(EXP_COLORS, EXP_COLORS == UNIQ_FACTORS[i], i + 1)
}

# remove unused variables
rm(DATA_FILES, CHARS_TO_REMOVE, COLUMNS_TO_READ, PKG_UPDATE)
```


Data summary:

Property             | Value
:---------           | :------
Genes                | `r dim(x)[1]`
Samples (conditions) | `r dim(x)[2]`
Unique factors       | `r toString(UNIQ_FACTORS)`
Colours              | `r toString(unique(EXP_COLORS))`


## Inspect raw data {#origSizeMDS}

Data will be inspected with three plots.

[START EXPANDIBLE]: #
`r EXPAND_bx`
Expand to read about <b>library size, MDS and PCA plots</b></summary>

* The first plot is for raw library sizes to see if they are balanced. 
`r NOTE_bx` 
Ideally, library sizes must be similar in all samples to avoid bias in type I error rate and FDR (_false discovery rate_, the type II error rate). We will see that `limma::lmFit()` after `limma::voom()` is the differential expression method better supporting unequal library sizes [@Law2014tt].
</div>
* The _Multidimensional scaling_ (**[MDS]**) scatterplot use a distance matrix to visualise the relative differences in transcriptional profile between the samples. It uses differences in expression profiles between different samples distances calculated on the `top = 500` genes with largest standard deviations across the samples in an iterative process. Replicates are expected to form clusters, that is, [the distance between replicates is shorter than with other samples](https://richardlent.github.io/post/multivariate-analysis-with-r/#multidimensional-scaling). Based on the `limma::plotMDS()` function [@Ritchie2015aa], [MDS] is the **best way to verify sample grouping** when replicates of different treatments are compared since it provides a type of unsupervised clustering of the samples. It is useful for examining how different are the profiles produced by different experimental factors and for identifying unexpected patterns.
* _Pincipal components analysis_ (**[PCA]**) is an easy and fast way to analyse data by a dimensional reduction of multivariate data, but based on the variance [@Jolliffe2016dp,@Ringner2008aa]. Since replicates should correlate, they are expected to form clusters in a [PCA] because the [principal components are linear combinations of the original variables that maximise the dispersion of samples](https://richardlent.github.io/post/multivariate-analysis-with-r/#principal-components-analysis). The representation shown is based on the RSeqFlow-defined function `PlotMyPCA()` that is a mere wrapper of the R function `stats::prcomp()`.

</details>
[END EXPANDIBLE]: #


```{r sizesMDSplots, fig.width=6, fig.height=5, out.width=c('33%', '33%', '33%'), fig.show='hold'} 
# grasp the names of samples to pinpoint them
COL_NAMES <- colnames(x)

# create a place at rigth (8) for the legend outside the graph
# structure of mar = c(bottom, left, top, right)
par(mar = c(4, 3, 3, 8), xpd = TRUE) 
# plot library sizes
barplot(x$samples$lib.size * 1e-6, 
        names = COL_NAMES, 
        col = EXP_COLORS,
        ylab = "Library size (millions)", 
        las = 2, 
        main = "Raw (original) data library sizes") 

# legend to indicate the correspondence of colours and factors in all plots
legend("topright", inset = c(-0.3, 0), # legend outside the plot 
       legend = as.vector(UNIQ_FACTORS), 
       fill = unique(EXP_COLORS),
       cex = 0.8)

# return to mar default values
par(mar = c(5, 4, 4, 2), xpd = TRUE)
# MDS: multi-dimensional scaling scatterplot
plotMDS(x, 
        col = EXP_COLORS, 
        main = "Raw sample similarities as MDS")


pca_sum <- PlotMyPCA(x$counts, "PCA of raw data", thisScale = FALSE)
```




## Save raw counts 

Raw counts are saved as _tab separated values_ (**.tsv**) only when several files were read to construct the expression table. In such a case, rows are gene codes and columns are experimental data. The user-defined function `SaveTSV()` will be used from now on to save data.


```{r datatable, eval=SAVE_RAW_DATA}
fileName <- SaveTSV(x$counts, "rawData-")
message("Raw counts table for all genes was saved in file", "\n", fileName)
```

