
## Heatmaps

[START EXPANDIBLE]: #
`r EXPAND_bx` Expand to read about <b>heatmaps</b></summary>

Heatmaps are another way to visualise [AHC] after [transforming data values to a color scale](https://statisticsglobe.com/heatmap-in-r). Its main advantage is that it allow us to simultaneously visualise clusters of samples and genes (rows, features), that is, how the genes (rows) change over an area (samples). Both columns and rows can be re-ordered according to the [AHC] result, putting similar observations close to each other. Visualising the data matrix in this way can help to find the variables (genes) that appear to be characteristic for each sample cluster. Ironically, the origin of the term "heat map" has nothing to do with temperature but it was invented to refer to graphical displays of real-time financial data.

</details> 
[END EXPANDIBLE]: #

1. Prepare data giving different colours for the corresponding groups of genes (rows) 

    ```{r geneclusters2}
    # HC_samples_ctf: hclust of samples, already determined 
    # HC_ctf_clust: hclust of samples already determined
    hc_colours <- sample(rainbow(final_k))
    mycolhc <- hc_colours[as.vector(groups_k)]
    ```


3. Plot the heatmap using the [AHC] clusters previously defined for rows (genes) using the `pheatmap::pheatmap()` function. It allows for more customisation and is more user-friendly than `heatmap()` and `heatmap.2()`. We will include the correspondence with clusters and experimental conditions, and show the heatmap with and withouth row scaling.

```{r pheatmap-complete, fig.align = 'center', results='hold', fig.width=6, fig.height=8, out.width = '59%', fig.show='hold'}
    # table with factors for the samples
    annot_columns <- data.frame(Sample = EXP_FACTORS,
                                row.names = colnames(m.putat.ctf.CV))
    
    # Create a list of colours for the samples
    my_colours <- list(Sample = setNames(EXP_COLORS, EXP_FACTORS))  # sample colouring defined by RSeqFlow
    
    # table with clusters for the genes
    annot_rows <- data.frame(AHC = paste0("AHC_", groups_k),
                             K_Means = paste0("KMeans_", groups_kmeans),
                             MClust = paste0("MClust_", groups_mclust),
                             row.names = rownames(m.putat.ctf.CV))
    
    # resulting heatmpap, more informative
    pheatmap(m.putat.ctf.CV, 
             color = HEATMAP_COLORS,
             scale = "row",                  # to have all rows (genes) in the same max-min scale
             border_color = "white",
             # customise columns
             show_colnames = TRUE,
             angle_col = 45,                 # rotate column names
             annotation_col = annot_columns, # add sample annotation
             cluster_cols = FALSE,           # if true, dendrogram and reorganisation of columns is produced
             clustering_distance_cols = d_ctf_samp,
             cutree_cols = k_samples,        # number of sample clusters calculated before
             # customise rows
             show_rownames = FALSE,
             annotation_row = annot_rows,    # add gene cluster annotation
             cluster_rows = TRUE,            # if false, no dendrogram and reorganisation of rows is produced
             clustering_distance_rows = d_ctf_genes, 
             cutree_rows = final_k,          # maximum number of clusters calculated before
             # colours for row and column annotations
             annotation_colors = my_colours,
             main = "Gene clusters respect to samples. Scaled by row")

    pheatmap(m.putat.ctf.CV, 
             color = HEATMAP_COLORS,
             scale = "none",                 # to have all rows (genes) in the same max-min scale
             border_color = NA,              # remove border, in this case is by default
             # customise names for columns and rows
             show_rownames = FALSE,
             show_colnames = TRUE,
             angle_col = 45,                 # rotate column names
             # customise annotations for columns and rows
             annotation_col = annot_columns, # add sample annotation
             annotation_row = annot_rows,    # add gene cluster annotation
             annotation_colors = my_colours,
             # customise clustering of columns and rows
             cluster_cols = TRUE,            # if true, dendrogram and reorganisation of columns is produced
             cluster_rows = TRUE,            # if false, no dendrogram and reorganisation of rows is produced
             clustering_distance_cols = d_ctf_samp,
             clustering_distance_rows = d_ctf_genes, 
             cutree_cols = k_samples,        # number of sample clusters calculated before
             cutree_rows = final_k,          # maximum number of clusters calculated before
             main = "Gene clusters respect to clustered samples. Not scaled")
```



## Aggregate matrix by factors

We will convert the expression of sample replicates in a single value by applying the `mean()` since it is more representative than me median when few replicates are available. The columns to be considered replicates of the same experimental condition was declared on `EXP_CONDITIONS` in `configure_wf-R`, and this was then converted into the new variable `EXP_FACTORS`.

Aggregation will be performed both with scaled expression and with [CPM] values. The new expression matrix aggregated by sample will then be used to plot cluster profiles, both from [AHC], _k_-means and [MBC].

1. Aggregate the **scaled matrix** `m.ctf.std` by sample using the mean. Transposition is required since the function will apply to columns.

    ```{r aggregate-scaled-samples}
    # aggregate by sample (columns). Transposition is required
    tmp <- t(aggregate(t(m.ctf.std), by = list(EXP_FACTORS), mean))
    colnames(tmp) <- tmp[1,] # add column names from the first row
    tmp <- tmp[-1, ] # remove the first row
    m_ctf_std_bysample <- apply(tmp, 2, as.numeric) # convert chars into numbers
    rownames(m_ctf_std_bysample) <- rownames(tmp)  # recuperate row names
    ```

2. For the same calculation in **[CPM]-transformed**, expression data from `x.filt.ctf` will be converted on [CPM] using `cpm()` and then filtered by gene names in `m.ctf.std`.

    ```{r aggregate-CPM-samples}
    # use the original set of counts to obtain the CPM of genes in clusters uniquely
    keep_genes <- rownames(x.filt.ctf) %in% rownames(m.ctf.std)
    m.ctf.cpm <- cpm(x.filt.ctf[keep_genes, ], log = FALSE)  # CPM (not log) only for genes in clusters
    # if log-CPM were calculated, the profile will be identical to scaled values.
    
    # aggregate samples
    tmp <- t(aggregate(t(m.ctf.cpm), by = list(EXP_FACTORS), mean))
    colnames(tmp) <- tmp[1,] # add column names from the first row
    tmp <- tmp[-1, ] # remove the first row
    m_ctf_cpm_bysample <- apply(tmp, 2, as.numeric) # convert chars into numbers
    rownames(m_ctf_cpm_bysample) <- rownames(tmp)  # recuperate row names
    ```

`r NOTE_bx` 
The resulting aggregated expression matrix are 

* `m_ctf_std_bysample` for scaled data
* `m_ctf_cpm_bysample` for CPM data
</div>

## Aggregate genes by cluster {.tabset .tabset-fade .tabset-pills}

Using the sample-aggregated matrices, genes will be aggregated by cluster using again the `mean()`. Error bars are based on the 95% confidence interval (`CI95`) using the standard deviation calculated by `sd()`, the number of genes in the cluster (stored in variable `NUM_genes_by_cluster`) and the critical _z_-value for the 95% (`z_CI`).

```{r z-for-CI95}
z_CI <- qnorm(p = 0.05/2, lower.tail = FALSE)  # z critical value for 95% confidence intervals

agg_list <- list()                             # create the list to gather aggregated data
```

`r NOTE_bx` 
The list `agg_list` will contain the aggregated genes by all [AHC] and _k_-means clusters for each aggregated set of replicates
</div>

### AHC-clusters {-}


1. Aggregate [AHC] clusters using scaled data:

    ```{r aggregateBy-AHCclusters-scaled}
    thisID <- "AHC_scaled"
    NUM_genes_by_cluster <- as.data.frame(table(groups_k))$Freq

    # aggregate the m_ctf_std_bysample matrix by clusters, using the mean, sd and CI95
    aggr_med <- aggregate(m_ctf_std_bysample, by = list(Cluster = groups_k), mean)
    tmp_sd <- aggregate(m_ctf_std_bysample, by = list(Cluster = groups_k), sd)
    # when there are clusters with only 1 value, sd returs NA. Change this for 0
    tmp_sd[is.na(tmp_sd)] <- 0
    aggr_CI95 <- cbind(tmp_sd[1], tmp_sd[, 2:ncol(tmp_sd)]*z_CI/sqrt(NUM_genes_by_cluster))
    aggr_CI95[is.na(aggr_CI95)] <- 0 # avoid crashing when NAs are present due to 1-member clusters
    
    # save aggregates in a list for further graphs
    agg_list[[thisID]]$med <- aggr_med
    agg_list[[thisID]]$CI95 <- aggr_CI95
    agg_list[[thisID]]$k <- final_k
    
    kable(cbind(Num_genes = NUM_genes_by_cluster, aggr_med), 
          align = "r", 
          digits = ROUND_dig, 
          caption = "Scaled, aggregated expression of clusters per experimental condition")
    ```


2. Aggregation of the same clusters using [CPM]s.

    ```{r aggregateBy-AHCclusters-CPM}
    thisID <- "AHC_cpm"

    # aggregate clusters by mean
    aggr_med <- aggregate(m_ctf_cpm_bysample, by = list(cluster = groups_k), mean)
    # standard deviation and 95% confidence interval
    tmp_sd <- aggregate(m_ctf_cpm_bysample, by = list(Cluster = groups_k), sd)
    # when there are clusters with only 1 value, sd returs NA. Change this for 0
    tmp_sd[is.na(tmp_sd)] <- 0
    aggr_CI95 <- cbind(tmp_sd[1], tmp_sd[, 2:ncol(tmp_sd)]*z_CI/sqrt(NUM_genes_by_cluster))
    aggr_CI95[is.na(aggr_CI95)] <- 0 # avoid crashing when NAs are present due to 1-member clusters
    
    # save aggregates in a list for further graphs
    agg_list[[thisID]]$med <- aggr_med
    agg_list[[thisID]]$CI95 <- aggr_CI95
    agg_list[[thisID]]$k <- final_k
    
    kable(cbind(Num_genes = NUM_genes_by_cluster, aggr_med), 
          align = "r", 
          digits = ROUND_dig, 
          caption = "Aggregated expression of clusters in CPMs per experimental condition")
    ```


### K-means clusters {-}

1. Aggregate _k_-means clusters using scaled data:

    ```{r aggregateBy-kmeans-scaled}
    thisID <- "kMeans_scaled"
    NUM_genes_by_cluster <- as.data.frame(table(groups_kmeans))$Freq

    # aggregate the new matrix by clusters, using the mean
    aggr_med <- aggregate(m_ctf_std_bysample, by = list(Cluster = groups_kmeans), mean)
    # standard deviation and 95% confidence interval
    tmp_sd <- aggregate(m_ctf_std_bysample, by = list(Cluster = groups_kmeans), sd)
    # when there are clusters with only 1 value, sd returs NA. Change this for 0
    tmp_sd[is.na(tmp_sd)] <- 0
    aggr_CI95 <- cbind(tmp_sd[1], tmp_sd[, 2:ncol(tmp_sd)]*z_CI/sqrt(NUM_genes_by_cluster))
    aggr_CI95[is.na(aggr_CI95)] <- 0 # avoid crashing when NAs are present due to one-member clusters
    
    # save aggregates in a list for further graphs
    agg_list[[thisID]]$med <- aggr_med
    agg_list[[thisID]]$CI95 <- aggr_CI95
    agg_list[[thisID]]$k <- kmeans_k
    
    kable(cbind(Num_genes = NUM_genes_by_cluster, aggr_med), 
          align = "r", 
          digits = ROUND_dig, 
          caption = "Scaled, aggregated expression of k-means clusters per experimental condition")
    ```

2. Aggregation of the same clusters using [CPM]s.

    ```{r aggregateBy-kmeans-CPM}
    thisID <- "kMeans_cpm"
    
    # aggregate clusters by mean
    aggr_med <- aggregate(m_ctf_cpm_bysample, by = list(cluster = groups_kmeans), mean)
    # standard deviation and 95% confidence interval
    tmp_sd <- aggregate(m_ctf_cpm_bysample, by = list(Cluster = groups_kmeans), sd)
    # when there are clusters with only 1 value, sd returs NA. Change this for 0
    tmp_sd[is.na(tmp_sd)] <- 0
    aggr_CI95 <- cbind(tmp_sd[1], tmp_sd[, 2:ncol(tmp_sd)]*z_CI/sqrt(NUM_genes_by_cluster))
    aggr_CI95[is.na(aggr_CI95)] <- 0 # avoid crashing when NAs are present due to one-member clusters
    
    # save aggregates in a list for further graphs
    agg_list[[thisID]]$med <- aggr_med
    agg_list[[thisID]]$CI95 <- aggr_CI95
    agg_list[[thisID]]$k <- kmeans_k
    
    kable(cbind(Num_genes = NUM_genes_by_cluster, aggr_med), 
          align = "r", 
          digits = ROUND_dig, 
          caption = "Aggregated expression of clusters in CPMs per experimental condition")
    ```


### MBC clusters {-}

1. Aggregation of [MBC] clusters using scaled data 

    ```{r aggregate-ByMclust-scale}
    thisID <- "MBC_scaled"
    n_genes_model_based <- table(groups_mclust)
    NUM_genes_by_cluster <- as.data.frame(n_genes_model_based)$Freq
    
    # aggregate clusters by mean
    aggr_med <- aggregate(m_ctf_std_bysample, by = list(Cluster = groups_mclust), mean)
    # standard deviation and 95% confidence interval
    tmp_sd <- aggregate(m_ctf_std_bysample, by = list(Cluster = groups_mclust), sd)
    # when there are clusters with only 1 value, sd returs NA. Change this for 0
    tmp_sd[is.na(tmp_sd)] <- 0
    aggr_CI95 <- cbind(tmp_sd[1], tmp_sd[, 2:ncol(tmp_sd)]*z_CI/sqrt(NUM_genes_by_cluster))
    aggr_CI95[is.na(aggr_CI95)] <- 0 # avoid crashing when NAs are present due to one-member clusters
    
    agg_list[[thisID]]$med <- aggr_med
    agg_list[[thisID]]$CI95 <- aggr_CI95
    agg_list[[thisID]]$k <- mclust_k
    
    kable(cbind(Num_genes = NUM_genes_by_cluster, aggr_med), 
          align = "r", 
          digits = ROUND_dig, 
          caption = "Scaled, aggregated expression of MBC clusters per experimental condition")
    
    thisID <- "MBC_cpm"
    ```


2. Aggregate [MBC] clusters using [CPM]s.

    ```{r aggregate-ByMclust-cpm}
    # aggregate clusters by mean
    aggr_med <- aggregate(m_ctf_cpm_bysample, by = list(cluster = groups_mclust), mean)
    # standard deviation and 95% confidence interval
    tmp_sd <- aggregate(m_ctf_cpm_bysample, by = list(Cluster = groups_mclust), sd)
    # when there are clusters with only 1 value, sd returs NA. Change this for 0
    tmp_sd[is.na(tmp_sd)] <- 0
    aggr_CI95 <- cbind(tmp_sd[1], tmp_sd[, 2:ncol(tmp_sd)]*z_CI/sqrt(NUM_genes_by_cluster))
    
    agg_list[[thisID]]$med <- aggr_med
    agg_list[[thisID]]$CI95 <- aggr_CI95
    agg_list[[thisID]]$k <- mclust_k
    
    rm(tmp_sd, z_CI)        # finally, remove temporal objects from RAM
    
    kable(cbind(Num_genes = NUM_genes_by_cluster, aggr_med), 
          align = "r", 
          digits = ROUND_dig, 
          caption = "Aggregated expression of clusters in CPMs per experimental condition")
    ```


## Cluster profiles

Let's see the expression profile of every cluster using the scaled expression and the expression in [CPM]. For statistical considerations, the plot with error bars based on the 95% confidence interval (`±CI95`) is also presented in the second plot column.

`r NOTE_bx` 
Errors bars are confidence intervals of the mean at 95% confidence). The profile of each gene in the cluster is represented in [section ***Gene profiles***](#gene-profiles) below.
</div>

```{r clustplotlog, fig.width=6, fig.height=4.5, out.width=c('50%', '50%')}
i <- 1
n_l <- length(agg_list)
# custom_lty <- 1:max(final_k, kmeans_k)  # for maximal number of clusters
niceCols <- brewer.pal(max(final_k, kmeans_k, mclust_k), "Dark2") # colours for the greatest k

for (i in 1:n_l) {
  case_name <- names(agg_list)[i]
  # tmp matrix without first row that contains cluster number after transposition
  plot_tmp <- t(agg_list[[i]]$med)[-1, ]
  thisTitle <- paste0("Expression course of ", case_name)
  MatPlot4Clusters(plot_tmp,
                   mainTitle = thisTitle,
                   myCols = niceCols)
  
  # print the same with error bars
  # based in https://stackoverflow.com/questions/40384055/add-error-bars-to-multiple-lines-to-show-standard-deviation-on-a-plot-in-r
  plot_tmp_sd <- t(agg_list[[i]]$CI95)[-1,]
  # margins of error per point
  low <- plot_tmp - plot_tmp_sd
  up <- plot_tmp + plot_tmp_sd

  x_pos <- 1:nrow(plot_tmp)            # x positions of each value
  par(mar = c(5, 4, 4, 2), xpd = TRUE) # define default figure size
  matplot(x_pos, plot_tmp, 
          main = paste0(thisTitle, " (±CI95)"),
          type = "b", 
          las = 2,                     # all labels perpendicular to axes
          pch = 0:25,                  # point symbols (up to 25 different types)
          col = niceCols,              # colours
          lty = 1:agg_list[[i]]$k,     # line type
          ylim = c(min(low), max(up)), # room for error bars within plot
          xaxt = "n",                  # remove default labels at x
          xlab = "Experimental condition",
          ylab = paste0("Expression median (", case_name, ")"))
  xx <- rep.int(x_pos, ncol(plot_tmp)) # recycle `x` for each column of plot_tmp
  repcols <- rep(niceCols, each = nrow(plot_tmp))  # recycle colours `col` for each row of `plot_tmp`
  # adding error bars using vectorization power of `arrow`
  arrows(xx, low, xx, up, col = repcols, angle = 90, length = 0.03, code = 3)
  axis(1,                           # below
       at = 1:nrow(plot_tmp),
       labels = rownames(plot_tmp),
       las = 2,
       cex.axis = 1)
}
```


## Gene profiles {.tabset .tabset-fade .tabset-pills}

While this may be aesthetically somewhat satisfactory, it is clear that the clusters are not homogenous as we might need them for biological interpretation. This is a general problem with clustering methods that fix the number of cluster centres either directly as in _k_-means or indirectly by cutting trees at a fixed level. It is also a problem with the data, where differences in absolute values might override separation into clusters that might better be defined in terms of relative values.

### AHC groups

1. Gene profiles as **scaled** values:

    ```{r maplots-AHCscaled, fig.width=4, fig.height=6, out.width=c('25%', '25%', '25%', '25%'), fig.show='hold'}
    # SCALED
    n_genes_group <- table(groups_k)
    i <- 1
    for (i in 1:final_k) {
      m2Plot <- t(m.ctf.std[groups_k == i, ])
      theTitle <- paste0(n_genes_group[i], " genes | AHC group ", i)
      print(PlotGeneProfiles(m2Plot, theTitle))
    }    
    ```


2. Gene profiles as **CPM** values

    ```{r maplots-AHCcpm, fig.width=4, fig.height=6, out.width=c('25%', '25%', '25%', '25%'), fig.show='hold'}
    # ORIGINAL COUNTS
    ylab_cpm <- "Count per million reads"
    i <- 1
    for (i in 1:final_k) {
      m2Plot <- t(m.ctf.cpm[groups_k == i, ])
      theTitle <- paste0(n_genes_group[i], " genes | AHC group ", i)
      print(PlotGeneProfiles(m2Plot, theTitle, ylab_cpm))
    }
    ```


### K-means groups

1. Gene profiles as **scaled** values.

    ```{r matplots-kmeansScaled, fig.width=4, fig.height=6, out.width=c('25%', '25%', '25%', '25%'), fig.show='hold'}
    i <- 1
    for (i in 1:kmeans_k) {
      m2Plot <- t(m.ctf.std[groups_kmeans == i, ])
      theTitle <- paste0(n_genes_group_kmeans[i], " genes | k-means group ", i)
      print(PlotGeneProfiles(m2Plot, theTitle))
    }    
    ```

2. Gene profiles as **CPM** values

    ```{r maplots-kmeans-cpm, fig.width=4, fig.height=6, out.width=c('25%', '25%', '25%', '25%'), fig.show='hold'}
    # ORIGINAL COUNTS
    i <- 1
    for (i in 1:kmeans_k) {
      m2Plot <- t(m.ctf.cpm[groups_kmeans == i, ])
      theTitle <- paste0(n_genes_group_kmeans[i], " genes | k-means group ", i)
      print(PlotGeneProfiles(m2Plot, theTitle, ylab_cpm))
    }
    ```


### MBC groups

1. Gene profiles as **scaled** values.

    ```{r matplots-MBC-Scaled, fig.width=4, fig.height=6, out.width=c('25%', '25%', '25%', '25%'), fig.show='hold'}
    i <- 1
    for (i in 1:mclust_k) {
      m2Plot <- t(m.ctf.std[groups_mclust == i, ])
      theTitle <- paste0(n_genes_model_based[i], " genes | MBC group ", i)
      print(PlotGeneProfiles(m2Plot, theTitle))
    }    
    ```

2. Gene profiles as **CPM** values

    ```{r maplots-MBC-cpm, fig.width=4, fig.height=6, out.width=c('25%', '25%', '25%', '25%'), fig.show='hold'}
        # ORIGINAL COUNTS
    i <- 1
    for (i in 1:mclust_k) {
      m2Plot <- t(m.ctf.cpm[groups_mclust == i, ])
      theTitle <- paste0(n_genes_model_based[i], " genes | MBC group ", i)
      print(PlotGeneProfiles(m2Plot, theTitle, ylab_cpm))
    }
    ```
