Here you can find interactive plots to inspect

- The **user-defined genes**, if any, in the `MY_IDs` parameter of the configuration file.
- Automatically selected [DEG]s in the `OUTSTANDING_GENE_LIST` variable, which merit a particular analysis by themselves.


## Selected by the user

```{r noOUTwarning1}
# Verify if there is any IDs defined in the configure_wf.R file
doID_PROFILES <- exists("MY_IDs")
if (doID_PROFILES) {
  OUTS_WARNING_MSG <- paste0("There are **", length(MY_IDs), " user IDs** to plot their profiles")
} else {
  OUTS_WARNING_MSG <- paste0(WARN_bx, " No user IDs were found in the configuration file; hence, nothing will be shown", "</div>")
}
```

`r OUTS_WARNING_MSG`.

```{r myIDs, fig.width=7, fig.height=5, out.width='90%', eval=doID_PROFILES, include=exists("MY_IDs")}
# using the filtered matrix in CTF
myIDs_2Plot <- t(x.filt.ctf[rownames(x.filt.ctf) %in% MY_IDs, ])
num_ID <- ncol(myIDs_2Plot)
theTitle <- paste0("Profiles for ", num_ID, " user-selected genes in the initially filtered matrix")
SelectedGeneProfilePlots(myIDs_2Plot,
                         aTitle = theTitle,
                         y_label = "CTF counts")

# using the scaled, filtered data
myIDs_2Plot <- t(m.ctf.std[rownames(m.ctf.std) %in% MY_IDs, ])
num_ID <- ncol(myIDs_2Plot)                     # number of myIDs found
theTitle <- paste0("Profiles for ", num_ID, " user-selected genes within clusterisable genes")
SelectedGeneProfilePlots(myIDs_2Plot,
                         aTitle = theTitle)
```



## Automatically selected {.tabset .tabset-fade .tabset-pill}

This set of genes are gathered in the `OUTSTANDING_GENE_LIST` variable constructed after clustering and networking [DEG]s. This list includes:

- Genes that are [DEG] in all contrasts (ubiquitous), including the `eBayes()` adjustment.
- Genes that are a significant hub (`kleinberg score >` `r MIN_KLEINBERG` and `deg > MIN_GENES_PER_CLUSTER`)

```{r summary}
selected_summary <- c(length(OUTSTANDING_GENE_LIST),
                      length(unlist(OUTSTANDING_GENE_LIST)),
                      length(unique(unlist(OUTSTANDING_GENE_LIST))))

# sort outstanding genes by decreasing frequency
outstand_sorted <- sort(table(unlist(OUTSTANDING_GENE_LIST)), decreasing = TRUE)

# genes qualified as outstanding in all conditions
outst_max <- max(outstand_sorted)
max_outs <- names(outstand_sorted[outstand_sorted == outst_max])

# select the most outstanding genes (above the median frequency of selection)
outst_median <- median(outstand_sorted)
most_outs <- names(outstand_sorted[outstand_sorted >= outst_median])
is_OUTS_EMPTY <- length(most_outs) < 1
if (is_OUTS_EMPTY) {
  most_outs <- max_outs
  OUTS_WARNING_MSG <- paste0(WARN_bx, " **No outstanding** transcript detected.", " </div>")
} else {
  OUTS_WARNING_MSG <- ""
}
```

`r OUTS_WARNING_MSG`

Summary: 

Feature | Amount
:---    | :---
Conditions with significant hubs      | `r selected_summary[1]`
Total IDs in the list                 | `r selected_summary[2]`
Unique ID in th list                  | `r selected_summary[3]`
The most frequent outstanding IDs     | `r length(most_outs)`
Outstanding IDs in nearly all samples | `r length(max_outs)`


### Reliability

The most reliable outstating genes must be selected in several analyses. Let's see if this is the case.

```{r prepSelectedGenes, fig.width=9, fig.height=5, fig.align='center', out.width='85%', results='hold', fig.show='hold'}
if (is_OUTS_EMPTY) {
  OUTS_WARNING_MSG <- paste0(WARN_bx, " There is **not any outstanding** transcript to plot", "</div>")
} else {
  OUTS_WARNING_MSG <- ""
  barplot(outstand_sorted, las = 2, main = "Outstanding genes", ylab = "Selection frequency")
}
```

`r OUTS_WARNING_MSG`.

### Profiles using the filtered dataset

Profiles of outstanding genes based on the expression in the filtered dataset `x.filt.ctf`, **if any**.

```{r max-outs-in-filtered, fig.width=7, fig.height=5, out.width='90%', eval=!(is_OUTS_EMPTY), include=!(is_OUTS_EMPTY)}
# Profiles of all outstanding genes, in `max_outs`
myIDs_2Plot <- t(x.filt.ctf[rownames(x.filt.ctf) %in% max_outs, ])
if (ncol(myIDs_2Plot) > 0) {
  theTitle <- paste0(length(max_outs), " gene profiles of the MOST outstanding genes")
  SelectedGeneProfilePlots(myIDs_2Plot,
                           aTitle = theTitle,
                         y_label = "CTF counts")
} else {
  warning("The MOST outstanding genes ", toString(max_outs), " are not among the filtered dataset of CTFs.")
}
```

Profiles of the most reliable outstanding genes in the filtered dataset, **if any**.

```{r most-outs-in-filtered, fig.width=7, fig.height=5, out.width='90%', eval=!(is_OUTS_EMPTY), include=!(is_OUTS_EMPTY)}
# Profiles the most frequent IDs, in 'most_outs', using the median as boundary
myIDs_2Plot <- t(x.filt.ctf[rownames(x.filt.ctf) %in% most_outs, ])
if (ncol(myIDs_2Plot) > 0) {
  theTitle <- paste0(length(most_outs), " (all) outstanding genes profiles")
  SelectedGeneProfilePlots(myIDs_2Plot,
                           aTitle = theTitle,
                         y_label = "CTF counts")
} else {
  warning("All the outstanding genes are not among the filtered dataset of CTFs.")
}
```


### Profiles using the clusterisable dataset

Here you can see the profiles of the most reliable outstanding genes, **if any**, using the expression data `m.ctf.std` for the clusterisable [DEG]s.

```{r max-outs-in-clust, fig.width=7, fig.height=5, out.width='90%', eval=!(is_OUTS_EMPTY), include=!(is_OUTS_EMPTY)}
# Profiles of all outstanding genes, in `max_outs`
myIDs_2Plot <- t(m.ctf.std[rownames(m.ctf.std) %in% max_outs, ])
if (ncol(myIDs_2Plot) > 0) {
  theTitle <- paste0(length(max_outs), " gene profiles of the MOST outstanding genes")
  SelectedGeneProfilePlots(myIDs_2Plot,
                         aTitle = theTitle)
} else {
  warning("The MOST outstanding genes ", toString(max_outs), " are not among clusterisable DEGs.")
}
```

Profiles of the most reliable outstanding genes in the cluterisable dataset, **if any**.

```{r most-outs-in-clust, fig.width=7, fig.height=5, out.width='90%', eval=!(is_OUTS_EMPTY), include=!(is_OUTS_EMPTY)}
# Profiles the most frequent IDs, in 'most_outs', using the median as boundary
myIDs_2Plot <- t(m.ctf.std[rownames(m.ctf.std) %in% most_outs, ])
if (ncol(myIDs_2Plot) > 0) {
  theTitle <- paste0(length(most_outs), " (all) outstanding genes profiles")
  SelectedGeneProfilePlots(myIDs_2Plot,
                           aTitle = theTitle)
} else {
  warning("All the outstanding genes are not among clusterisable DEGs.")
}
```




## Save outstanding genes

```{r noOUTwarning2}
# Verify if there are outstanding genes in the analysis
if (is_OUTS_EMPTY) {
  OUTS_WARNING_MSG <- paste0(WARN_bx, " **No OutstandingGenes-... file will be saved:** there is no ID to save", "</div>")
} else {
  OUTS_WARNING_MSG <- ""
}
```

Saving the analysis that produced outstanding genes as well as the list of gene IDs for each condition (as well as the user defined IDs, if any), provided that you have any outstanding transcript to save. `r OUTS_WARNING_MSG`.

```{r saveOutstanding, eval=!is_OUTS_EMPTY, include=!is_OUTS_EMPTY}
# Add the user defined IDs, if any
if (doID_PROFILES) OUTSTANDING_GENE_LIST[["User_defined"]] <- MY_IDs

m <- as.matrix(unlist2(OUTSTANDING_GENE_LIST))   # convert the list into a long matrix
colnames(m) <- "GeneID"                          # put a meaningful name to the gene column
fileName <- SaveTSV(m, "OutstandingGenes-")      # save outstanding genes

message("Outstanding genes were saved in file", "\n", fileName)
```

