Profiles of genes that have been selected by at least one of this criteria:

1. Automatically selected:
    * Being a DEG in all contrasts (ubiquitous)
    * Being a significant hub (`kleinberg score >` `r MIN_KLEINBERG` and `deg > MIN_GENES_PER_CLUSTER`)
3. Defined by the user: in `MY_IDs` at `configure_wf.R` file

Those genes merit a particular analysis by themselves.

## Automatically selected

Preparing data for representations:

```{r prepSelectedGenes, fig.width=8, fig.height=6, fig.align='center', out.width=c('60%')}
selected_summary <- c(length(OUTSTANDING_GENE_LIST),
                      length(unlist(OUTSTANDING_GENE_LIST)),
                      length(unique(unlist(OUTSTANDING_GENE_LIST))))
outstand_sorted <- sort(table(unlist(OUTSTANDING_GENE_LIST)), decreasing = TRUE)
barplot(outstand_sorted, las = 2, main = "Outstanding genes", ylab = "Selection frequency")

# select the most outstanding genes (above the median frequency of selection)
outst_median <- median(outstand_sorted)
most_outs <- names(outstand_sorted[outstand_sorted > outst_median])
```

Summary: 

Feature | Amount
:---    | :---
Conditions with significant hubs | `r selected_summary[1]`
Total ID selected                | `r selected_summary[2]`
Unique ID from the total         | `r selected_summary[3]`
The most outstanding IDs         | `r length(most_outs)`

Let's show the gene profiles of the most frequently selected genes at left and the others at right.

```{r selgeneprof, fig.width=5, fig.height=4, out.width=c('50%', '50%'), fig.show='hold'}
# represent the most frequent and the others, using the median as boundary
myIDs_2Plot <- t(m.ctf.std[rownames(m.ctf.std) %in% most_outs, ])
theTitle <- "Gene profiles of most outstanding genes"
SelectedGeneProfilePlots(myIDs_2Plot,
                         aTitle = theTitle)

other_outs <- names(outstand_sorted[outstand_sorted <= outst_median])
myIDs_2Plot <- t(m.ctf.std[rownames(m.ctf.std) %in% other_outs, ])
theTitle <- "Profiles for other selected genes"
SelectedGeneProfilePlots(myIDs_2Plot,
                         aTitle = theTitle)

```

## Selected by the user

Some genes may be selected by the user during this analysis, and can be seen here, provided that `MY_IDs` is not empty in `configure_wf.R`.

```{r myIDs, fig.width=5, fig.height=4, out.width=c('50%', '50%'), fig.show='hold', eval=exists("MY_IDs"), include=exists("MY_IDs")}
# Add user selected genes to the OUTSTANDING variable for further saving
OUTSTANDING_GENE_LIST[["User_defined"]] <- MY_IDs

# using the scaled, filtered data
myIDs_2Plot <- t(m.ctf.std[rownames(m.ctf.std) %in% MY_IDs, ])
theTitle <- "Profiles for user-selected genes in scaled"
SelectedGeneProfilePlots(myIDs_2Plot,
                         aTitle = theTitle)

# scaling the CTF-normalised counts to include previously filtered-out genes
m.filt.ctf_std <- scale(x.filt.ctf)
myIDs_2Plot <- t(m.filt.ctf_std[rownames(m.filt.ctf_std) %in% MY_IDs, ])
theTitle <- "Profiles for user-selected genes as CTF-scale counts"
SelectedGeneProfilePlots(myIDs_2Plot,
                         aTitle = theTitle,
                         y_label = "CTF-scaled counts")
```



## Save outstanding genes

Saving the analysis that produced outstanding genes as well as the list of gene IDs for each condition.

```{r saveOutstanding}
# convert the list into a long matrix
m <- as.matrix(unlist2(OUTSTANDING_GENE_LIST))
# put a meaningful name to the gene column
colnames(m) <- "GeneID"
fileName <- SaveTSV(m, "OutstandingGenes-")
message("Outstanding genes were saved in file", "\n", fileName)
```

