## Removing noisy genes

Unexpressed genes are those with `counts == 0` in all samples (conditions, e.g. columns) and should be removed since they have no biological interest and only introduce noise in statistics.

```{r filter1, results='hold'}
# Looking for genes expressed in all samples (TRUE)
none_is0 <- table(rowSums(x$counts == 0) == 0)

# Counting genes not expressed in any sample (TRUE)
COL_NUMBER <- length(COL_NAMES)
all_are0 <- table(rowSums(x$counts == 0) == COL_NUMBER)

# other counts
total_genes <- dim(x)[1] # or nrow(x$counts)
at_least_1 <- total_genes - none_is0[[2]] - all_are0[[2]]
```

We can remove at the same time those unexpressed genes together with any **noisy genes** (genes with unreliable expression because they have low mapping counts [_underexpression_] in most samples). We will consider a minimal reliable expression when genes are expressed in at least one set of replicates.

> **IMPORTANT:** The analysis cannot rely on `counts` since they depend on library sizes. [CPM] (_counts per million_) will be used instead of counts:

    ```{r x.cpm}
    # CMP for the x object containing raw data
    x.cpm <- cpm(x)
    ```

Removing `CPM > 0` will remove genes with no expression in all samples. Increasing the threshold to `MIN_CPM` value defined in `configure_wf.R` will remove also the low-expressed genes.

The minimal number of samples with `cpm > MIN_CPM` is the lowest number of replicates as defined in `EXP_FACTORS`.

```{r numRep}
# calculate the minimal number of samples for cpm > MIN_CPM
MIN_REPL <- min(table(EXP_FACTORS))

# filtering by CPM
keep.exprs <- rowSums(x.cpm > MIN_CPM) >= MIN_REPL
filteredOK <- table(keep.exprs) # see in TRUE how many genes pass the filter
x.filt <- x[keep.exprs, , keep.lib.sizes = FALSE]
```

The final balance is the following:

Genes | Amount  | Percent (%)
:---- | ----:   | ---:
Total                                   | `r total_genes`       | 100
Expressed in all samples                | `r none_is0[[2]]`  | `r round(100 * none_is0[[2]]/total_genes)`
Not expressed in at least 1 sample      | `r at_least_1`     | `r round(100 * at_least_1/total_genes)`
Never expressed                         | `r all_are0[[2]]`  | `r round(100 * all_are0[[2]]/total_genes)`
Tentetively useful                      | `r all_are0[[1]]`  | `r round(100 * all_are0[[1]]/total_genes)`
Really useful after `CMP >` `r MIN_CPM` | `r filteredOK[[2]]`| `r round(100 * filteredOK[2][1]/total_genes)`



## Data without noise {.tabset .tabset-fade .tabset-pills}

### Density plots {-}

[START EXPANDIBLE]: #
`r EXPAND_bx` 
Expand to read about the <b>density plot</b></summary>

The density of logCPM (`lcpm`) values will be plotted for each sample, where

* _dotted_ vertical line at `CPM = 0` (equivalent to `logCPM = 1`) will pinpoint the minimum logCPM threshold;
* _dashed_ vertical line will mark the main zone of gene removal. 

These reference lines are calculated as follows:

```{r cutoffs}
# To define the same range of density for all data
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
# a cutoff line to sepparate useful from useless genes
genes.cutoff <- log2(10/M + 2/L)
# remove needless variables
rm(L, M)
```

</details> 
[END EXPANDIBLE]: #


Density plots are produced using `RColorBrewer` within local function `PlotGeneDensity()` to individualise samples.

```{r densityplots, fig.width=4, fig.height=5, out.width=c('50%', '50%'), fig.show='hold'}
# calculation of CPM and logCPM
# for raw data (x.cpm was already calculated)
x.lcpm <- cpm(x, log = TRUE)
# for filtered data
x.filt.lcpm <- cpm(x.filt, log = TRUE)
x.filt.cpm <- cpm(x.filt)

# density plots of raw data and filtered data
PlotGeneDensity(x.lcpm, genes.cutoff, "Raw data")
PlotGeneDensity(x.filt.lcpm, genes.cutoff, "After filtering")

# remove needless variables
rm(none_is0, all_are0, filteredOK, keep.exprs) # from previous chunk
rm(genes.cutoff, x.lcpm, x.filt.lcpm)
```

> **IMPORTANT:** After filtering, density lines should be nearly identical following a gauss-shaped profile. Peak at very low logCPM should disappear.


***

### Inspect filtered data {#filterdSizeMDSPCA}

Let`s see the new library sizes and sample grouping in [MDS] and [PCA] plots to compare with previous results with [raw data](#origSizeMDS).


```{r sizesMDSplotsFiltered, fig.width=6, fig.height=5, out.width=c('33%', '33%', '33%'), fig.show='hold'}
# Reconstruct libsizes after filtering.
# These new, effective library sizes will be used  in all downsteam analyses
x.filt$samples$lib.size <- colSums(x.filt$counts)

# create a place at rigth (8) for the legend outside the graph
# structure of mar = c(bottom, left, top, right)
par(mar = c(4, 3, 3, 8), xpd = TRUE) 

barplot(x.filt$samples$lib.size * 1e-6, 
        names = COL_NAMES, 
        col = EXP_COLORS,
        ylab = "Library size (millions)", 
        las = 2, 
        main = "Library sizes after filtering") 

# legend to indicate the correspondence of colours and factors in all plots
legend("topright", inset = c(-0.3, 0), # legend outside the plot 
       legend = as.vector(UNIQ_FACTORS), 
       fill = unique(EXP_COLORS),
       cex = 0.8)

# return to mar default values
par(mar = c(5, 4, 4, 2), xpd = TRUE)

plotMDS(x.filt, 
        col = EXP_COLORS, 
        main = "MDS relative similarities after filtering")

pca_sum <- PlotMyPCA(x.filt$counts, "PCA of filtered data")
```

> **IMPORTANT:** Library sizes appear to be the same than in [previous plot](#origSizeMDS) since only genes with very few counts were removed.

`r NOTE_bx` Samples should group similar to [previous MDS and PCA plots](#origSizeMDS) or **even better**, since filtering must remove noisy data and improve sample clustering.</div>

`r WARN_bx` If inferior sample grouping, you have a problem with your samples. Take results with caution.</div>



***


## Save filtered counts 

Gene counts after filtering are saved in a **.tsv** (_tab separated values_) file where gene codes are in rows and experimental data in columns.

```{r datatablefiltered}
# SaveTSV() is difined in functions_wr.R
fileName <- SaveTSV(x.filt$counts, "filteredData-")
message(fileName, "\n", "contains filtered gene counts")
```
