Let's mathematically define the experiment considering that we have a control (`CTRL`) vs. treatment (`TREAT`). In the present version, we have included two more treatments (`TREAT2` and `TREAT3`) for the time course data.

1. Convert the experimental design in a matrix where rows are samples and columns are each experimental condition/treatment (derive from `EXP_FACTORS` configuration).
    ```{r design}
    # define el diseño en función de los grupos definidos al cargar los datos
    design <- model.matrix(~0+EXP_FACTORS)
    # retiramos ahora la palabra "EXP_FACTORS" de cada columna
    colnames(design) <- gsub("EXP_FACTORS", "", colnames(design))
    # Mostramos el diseño experimental
    kable(design, align = "c", caption = "Table describing the experimental design")
    ```


2. The contrasts are automatically obtained from the `CONTRAST` list defined in `configure_wf.R` file that was internally converted into `allContrasts`. **Pay attention** since this es very tricky. Combined with the experimental `design`, it enables all comparisons.
    ```{r contrasts}
    # allContrasts was calculated from `CONTRAST` in execute_wf.R file
    # defining contrast matrix for all comparisons
    # (the contrast matrix corresponding to specified contrasts of a set of parameters)
    contr_matrix <- makeContrasts(contrasts = allContrasts, levels = colnames(design))
    kable(contr_matrix, align = "c", caption = "Table of contrast that will be analysed")
    ```
